---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Read in the dataset of steps taken by a person for 2 months, measured in 5 minute intervals.  
The dataset is stored in activity.zip included in this repository.  
```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv")
str(activity)
```
The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken, the last two digits mean the minutes of the hour
    the first digits the hour, so 2355 means the interval from 23:55 to 24:00 hour.
    
## What is mean total number of steps taken per day?
To calculate the total number of steps taken per day we make a dataset where the steps are summarised
over the different dates. The missing values are ignored for the moment.  
```{r}
StepsPerDay <- aggregate(steps ~ date, data=activity, sum)
str(StepsPerDay)
```
Then we make it visible with a histogram of the total number of steps taken each day.  
```{r}
hist(StepsPerDay$steps, 
     main = "Histogram of total number of steps per day",
     xlab = "Total steps per day")
```

Finally we calculate and report the mean and median of the total number of steps taken each day. 
```{r}
mean(StepsPerDay$steps, na.rm=TRUE)
median(StepsPerDay$steps, na.rm=TRUE)
```

```{r mean, echo=FALSE, results = 'hide'}
meandata <- mean(StepsPerDay$steps, na.rm=TRUE)
```

```{r median, echo =FALSE, results='hide'}
mediandata <- median(StepsPerDay$steps, na.rm=TRUE)
```
So the mean `r meandata` and the median `r mediandata` are quite close together.  

## What is the average daily activity pattern?
To see what is the avarage daily pattern we first have to calculate the average 
per time interval over all the days in the dataset.
```{r}
StepsPerInterval <- aggregate(steps ~ interval, data = activity, mean, na.rm=TRUE)
str(StepsPerInterval)
```
Which we can now show in a nice plot.
```{r}
plot(StepsPerInterval$interval, StepsPerInterval$steps, type = "l",
     main = "Average daily activity pattern",
     xlab = "5-minute time interval",
     ylab = "Number of steps taken")
```

It seems our test subject does some daily excersise somewhere between the 5:00 and 10:00 a.m.  
Which 5-minute interval contains the maximum numbers of steps?
```{r}
StepsPerInterval[StepsPerInterval$steps == max(StepsPerInterval$steps), ]
```
It seem the test subject is on average taking the most steps around 8:35 am.  

## Imputing missing values
Perhaps we have introduced some bias in this research by ignoring the missing values which are in the dataset activity.
```{r}
head(activity)
```

Let's see how many missing values there are in the dataset **activity**.
```{r}
missing <- sum(!complete.cases(activity))
total <- nrow(activity)
percentage <- round(missing/total*100, digits = 1)
```
That is `r missing` missing values on `r total` observations, which is `r percentage` %.  
That is quite large, so let's see what happens when we impute the data.  
First select the complete and incomplete cases.
From the incomplete cases we drop the first column since there are only NA's in it.
```{r}
complete <- activity[complete.cases(activity),]
incomplete <- activity[!complete.cases(activity),][-c(1)]
head(complete)
head(incomplete)
```

Then we use the **StepsPerInterval** dataset with the mean value per interval
to give a (mean) value to the incomplete cases.
```{r}
completed <- merge(incomplete, StepsPerInterval, by = "interval")
head(completed)
```
Finally we create a new dataset **activityImputed** which contains mean 
values for the missing data. Since the order of the data is altered 
by the merge and the pasting together of the data, we sort the final
dataset again in the original order of **activity**.
```{r}
activityImputed <- rbind(complete, completed)
activityImputed <- activityImputed[order(activityImputed$date,activityImputed$interval),]
head(activityImputed)
```
Now we can analyse this imputed dataset and see if there is any difference.  
First calculate the total number of steps taken per day, then make a histogram of it.  
```{r}
StepsPerDayImp <- aggregate(steps ~ date, data = activityImputed, sum)
hist(StepsPerDayImp$steps,
     main = "Histogram of total number of steps per day with imputed data",
     xlab = "Total steps per day")
```

The original mean of the total steps taken per day was `r meandata`.
The mean of the imputed data is:
````{r}
mean(StepsPerDayImp$steps, na.rm=TRUE)
```
Because we imputed the data with the mean value, the mean value will not alter.  

The original median of the total steps taken per day was `r mediandata`. The median of the imputed data is:
```{r}
median(StepsPerDayImp$steps, na.rm=TRUE)
```
Because we added values the median will move to the mean value.  
This also shows in the histogram. The frequency of the interval where the mean value is, is higher.  

## Are there differences in activity patterns between weekdays and weekends?
Perhaps there is different behaviour in the weekend than on workdays.  
To investigate that we need to create a new factor variable in the 
dataset **activityImputed** with two levels: "weekend" and "weekday". 
(Because my dutch locale can not be changed it says "zaterdag" for "Saturday" and "zondag" for "Sunday")
```{r, results="hide"}
day <- weekdays(as.Date(activityImputed$date))
day[day %in% c("zaterdag","zondag")] <- "weekend"
day[!(day %in% c("weekend"))] <- "weekday"
factor(day)
activityImputed <- cbind(activityImputed,day)
```
```{r}
head(activityImputed)
```
Then we calculate the average steps per time interval again, but now also per type of day.
```{r}
StepsPerIntervalImp <- aggregate(steps ~ interval+day, data = activityImputed, mean)
head(StepsPerIntervalImp)
tail(StepsPerIntervalImp)
str(StepsPerIntervalImp)
```
So the daily patterns change a little in weekends as compaired to weekdays as can be seen in this plot.
```{r}
lattice::xyplot(steps ~ interval | day, data = StepsPerIntervalImp, layout = c(1,2), type="l")
```

In weekends the testsubject moves a little less between 5:00 and 10:00 and a little more and longer for the rest of the day.